---
title: "Create cross-matched SignatureAnalyzer and SigProfiler synthetic data to explore handling of flat signatures"
author: "Steve Rozen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create cross-matched SignatureAnalyzer and SigProfiler synthetic data to explore handling of flat signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Conceptual challenges arise when comparing extraction of signatures between
different categorizations of mutation types.  

For example, in PCAWG7, SigProfiler reference signatures are based on
extraction of 96-channel signatures (i.e. single base substitutions -- SBS or SNS),
while SignatureAnalyzer reference signatures  are based on extraction of
"COMPOSITE" signatures consisting of SBS in pentanucleotide context plus
double base substitutions (DBSs) and indels (IDs).

A second challenge in generating "realistic" synthetic data is deciding
how many signatures operate in a given tumor time. Here we address
this challenge by generating different synthetic data sets, one
based on the larger number of signatures as extracted by
SignatureAnalyzer, and one based on the smaller number of signatures
extracted by SigProfiler.

In the end we take the cross product of the two choices, that is,
```
 {96-channel, COMPOSITE} X {SignatureAnalyzer-attributions, SigProfiler-attributions}
```

## Required libraries

```{r library}
library(SynSig)
```

## Version 1, Kidney-RCC and ovarian adenocarcinoma to explore SBS3, SBS5, and SBS40

```{r}
SetNewOutDir("tmp.syn.3.5.40.rcc.and.ovary")
```

```{r}
set.seed(191905)
num.syn.tumors <- 500

rcc.info <- 
  SAAndSPSynDataOneCAType(
    sa.no.hyper.real.exposures,
    sp.no.hyper.real.exposures,
    ca.type = "Kidney-RCC",
    num.syn.tumors,
    file.prefix = "RCC")
```

```{r}
ovary.info <- 
  SAAndSPSynDataOneCAType(
    sa.no.hyper.real.exposures,
    sp.no.hyper.real.exposures,
    ca.type = "Ovary-AdenoCA",
    num.syn.tumors,
    file.prefix = "OVA")
```

Combine the RCC and ovarian adenocarcinoma synthetic exposures.

```{r} 
sa.rcc.ova.syn.exp <-
  MergeExposures(list(rcc.info$sa.syn.exp, ovary.info$sa.syn.exp))

sp.rcc.ova.syn.exp <-
  MergeExposures(list(rcc.info$sp.syn.exp, ovary.info$sp.syn.exp))

# We might want the exposures later to understand the 
# characteristics of the synthetic data.
WriteExposure(sa.rcc.ova.syn.exp, OutDir("sa.exposure.csv"))
WriteExposure(sp.rcc.ova.syn.exp, OutDir("sp.exposure.csv"))

```

### Generate synthetic catalogs for RCC combined with ovary adenocarcinoma

#### Create synthetic mutational spectra catalogs based on SignatureAnalyzer attributions

```{r}
CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sa.rcc.ova.syn.exp,
  "sa.sa.COMPOSITE",
  WriteCatCOMPOSITE)
```

```{r}
CreateAndWriteCatalog(
  sa.96.sigs,
  sa.rcc.ova.syn.exp,
  "sa.sa.96",
  ICAMS::WriteCatSNS96)

```

#### Create synthetic catalogs based on SigProfiler attributions

Create the catalogs. First we need the matching between SigProfiler
and SignatureAnalyzers signatures.

```{r}
sp.sa.map.info <-
  MapSPToSASignatureNamesInExposure(sp.rcc.ova.syn.exp)

CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sp.sa.map.info$exp2,
  "sp.sa.COMPOSITE",
  WriteCatCOMPOSITE)
```

### The mapping from SP signatures to SA signatures

```{r}
knitr::kable(
  sp.sa.map.info$sp.to.sa.sig.match, 
  caption = 'Best matches from SP signatures to SA signatures',
  digits = 4
)
```

```{r}
CreateAndWriteCatalog(
  sp.sigs,
  sp.rcc.ova.syn.exp,
  "sp.sp",
  ICAMS::WriteCatSNS96)

```

## Version 2: An "abstract" synthetic data set with only SBS3, 5, and 40

### Pull together the parameters

```{r}
x.sp.parms <-
  cbind(rcc.info$sp.parms[ , c("SBS5", "SBS40")],
        ovary.info$sp.parms[ , "SBS3", drop = FALSE])

# Find mapping from SBS3, SBS5, and SBS40 to SignatureAnalyzer signatures
# assigned to these tumor types

MatchSigs1Direction(
  sp.sigs[ , "SBS5", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(rcc.info$sa.syn.exp))])

MatchSigs1Direction(
  sp.sigs[ , "SBS40", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(rcc.info$sa.syn.exp))])

MatchSigs1Direction(
  sp.sigs[ , "SBS3", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(ovary.info$sa.syn.exp))])


# Both BI..SBS3 and BI..SBS39 are in every ovarian; we select BI..SBS3

x.sa.parms <-
  cbind(rcc.info$sa.parms[ , c("BI_COMPOSITE_SBS5_P",
                               "BI_COMPOSITE_SBS40_P")],
        ovary.info$sa.parms[ , "BI_COMPOSITE_SBS3_P", drop = FALSE])

knitr::kable(
  x.sa.parms, 
  caption = 'SignatureAnalyzer parameters',
  digits = 4
)

knitr::kable(
  x.sp.parms, 
  caption = 'SigProfiler parameters',
  digits = 4
)
```


```{r}
SetNewOutDir("tmp.syn.3.5.40.abst")
set.seed(44)
num.syn.tumors <- 1000

sp.abst.info <-
  GenerateSynAbstract(
    x.sp.parms, num.syn.tumors, "sp", "SP.Syn.Abst")

sa.abst.info <-
  GenerateSynAbstract(
    x.sa.parms,
    num.syn.tumors, "sa", "SA.Syn.Abst")

```

#### Generate and write SignatureAnalyzer "abstract" 3, 5, 40 catalogs

```{r}
CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sa.abst.info$syn.exp,
  "sa.sa.COMPOSITE",
  WriteCatCOMPOSITE)

CreateAndWriteCatalog(
  sa.96.sigs, 
  sa.abst.info$syn.exp,
  "sa.sa.96",
  ICAMS::WriteCatSNS96)

```

#### Generate and write SigProfiler "abstract" 3, 5, 40 catalogs

We need to adjust the signature names in the exposures
so they match the signature names in \code{sa.COMPOSITE.sigs}.

```{r}
tmp.exp <- sp.abst.info$syn.exp
rownames(tmp.exp) <- rownames(sa.abst.info$syn.exp)

CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  tmp.exp,
  "sp.sa.COMPOSITE",
  WriteCatCOMPOSITE)

CreateAndWriteCatalog(
  sp.sigs, 
  sp.abst.info$syn.exp,
  "sp.sp",
  ICAMS::WriteCatSNS96)

```

## Check results

Set <tt>unlink = FALSE</tt> if you want to keep the results.
Return values are `"ok"` on success.

```{r}
Diff4SynDataSets("syn.3.5.40.rcc.and.ovary", unlink = TRUE)

Diff4SynDataSets("syn.3.5.40.abst", unlink = TRUE)
```
