---
title: "Create cross-matched SignatureAnalyzer and SigProfiler synthetic data to explore handling of overlapping / interfering signatures"
author: "Steve Rozen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create cross-matched SignatureAnalyzer and SigProfiler synthetic data to explore handling overlapping / interfering signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview and rationale

Checking for interference between C > T mutations in 
APOBEC (SBS2) and UV (SBS7a,b).

## Required libraries

```{r library}
library(SynSig)
```

## Data set 1, bladder and skin melanoma to explore SBS2 and SBS7a,b

500 synthetic bladder transitional cell carcinomas (high in SBS2 and SBS13) 
and 500 synthetic skin melanomas (high in SBS7a,b,c,d).  The potential interference is between 
SBS2 (mainly C > T) and SBS7a,b (mainly C > T).

```{r}
SetNewOutDir("tmp.syn.2.7a.7b.bladder.and.melanoma")
```

```{r}
set.seed(191906)
num.syn.tumors <- 500

bladder.info <- 
  SAAndSPSynDataOneCAType(
    sa.no.hyper.real.exposures,
    sp.no.hyper.real.exposures,
    ca.type = "Bladder-TCC",
    num.syn.tumors,
    file.prefix = "BLADDER")
```

```{r}
skin.info <- 
  SAAndSPSynDataOneCAType(
    sa.all.real.exposures,
    sp.all.real.exposures,
    ca.type = "Skin-Melanoma",
    num.syn.tumors,
    file.prefix = "SKIN")
```

Combine the bladder and skin melanoma synthetic exposures.

```{r} 
# debug(MergeExposures)
sa.bladder.skin.syn.exp <-
  MergeExposures(list(bladder.info$sa.syn.exp, skin.info$sa.syn.exp))

sp.bladder.skin.syn.exp <-
  MergeExposures(list(bladder.info$sp.syn.exp, skin.info$sp.syn.exp))

# We will need the exposures later when evaluating the attributed signatures
WriteExposure(sa.bladder.skin.syn.exp, OutDir("sa.exposure.csv"))
WriteExposure(sp.bladder.skin.syn.exp, OutDir("sp.exposure.csv"))

```

### Generate synthetic catalogs for bladder combined with skin melanoma

#### Create synthetic mutational spectra catalogs based on SignatureAnalyzer attributions

```{r}
CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sa.bladder.skin.syn.exp,
  "sa.sa.COMPOSITE",
  WriteCatCOMPOSITE)


CreateAndWriteCatalog(
  sa.96.sigs,
  sa.bladder.skin.syn.exp,
  "sa.sa.96",
  WriteCatSNS96)

```

#### Create synthetic catalogs based on SigProfiler attributions

Create the catalogs. First we need the matching between SigProfiler
and SignatureAnalyzer signatures.

```{r}
sp.sa.map.info <-
  MapSPToSASignatureNamesInExposure(sp.bladder.skin.syn.exp)

CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sp.sa.map.info$exp2,
  "sp.sa.COMPOSITE",
  WriteCatCOMPOSITE)
```

```{r}
knitr::kable(
  sp.sa.map.info$sp.to.sa.sig.match, 
  caption = 'Best matches from SP signatures to SA signatures',
  digits = 4)
```

```{r}
CreateAndWriteCatalog(
  sp.sigs,
  sp.bladder.skin.syn.exp,
  "sp.sp",
  WriteCatSNS96)

```

### Copy in the scripts

```{r}
AddAllScripts()
```

## Version 2: "abstract" synthetic data set from only SBS2, SBS7a, and SBS7b 

1000 synthetic tumors composed from SBS2 and SBS7a,b.
Mutation load distributions 
drawn from bladder transitional cell carcinoma (SBS2) and
from skin melanoma (SBS7a,b).
Most spectra contain both signatures.
The potential interference is between SBS2 (mainly C > T) 
and SBS7a,b (mainly C > T).

### Pull together the parameters

```{r}
x.sp.parms <-
  cbind(bladder.info$sp.parms[ , "SBS2", drop = FALSE],
        skin.info$sp.parms[ , c("SBS7a", "SBS7b")])

# Find mapping from SBS2, SBS7a, and SBS7b to SignatureAnalyzer signatures
# assigned to these tumor types

MatchSigs1Direction(
  sp.sigs[ , "SBS2", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(bladder.info$sa.syn.exp))])

MatchSigs1Direction(
  sp.sigs[ , "SBS7a", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(skin.info$sa.syn.exp))])

MatchSigs1Direction(
  sp.sigs[ , "SBS7b", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(skin.info$sa.syn.exp))])


# We select BI..SBS2, BI....SBS7a, BI.....SBS7b from
# the SignatureAnalyzer parameters.

x.sa.parms <-
  cbind(bladder.info$sa.parms[ , "BI_COMPOSITE_SBS2_P", drop = FALSE],
        skin.info$sa.parms[ , c("BI_COMPOSITE_SBS7a_S", 
                                "BI_COMPOSITE_SBS7b_S")])

knitr::kable(
  x.sa.parms, 
  caption = 'SignatureAnalyzer parameters',
  digits = 4
)

knitr::kable(
  x.sp.parms, 
  caption = 'SigProfiler parameters',
  digits = 4
)
```

### Generate synthetic exposures

```{r}
SetNewOutDir("tmp.syn.2.7a.7b.abst")
set.seed(55)
num.syn.tumors <- 1000
```

```{r SP-abs-syn-exp}
sp.abst.info <-
  GenerateSynAbstract(
    x.sp.parms, num.syn.tumors, "sp", "SP.Syn.Abst")
```

```{r SA-abs-syn-exp}
# debug(present.sigs)
# debug(GenerateSyntheticExposures)
sa.abst.info <-
  GenerateSynAbstract(
    x.sa.parms, num.syn.tumors, "sa", "SA.Syn.Abst")

```

### Generate and write SignatureAnalyzer "abstract" 2, 7a, 7b catalogs

```{r SA-abs-syn-cat}
CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sa.abst.info$syn.exp,
  "sa.sa.COMPOSITE",
  WriteCatCOMPOSITE)

CreateAndWriteCatalog(
  sa.96.sigs, 
  sa.abst.info$syn.exp,
  "sa.sa.96",
  WriteCatSNS96)

```

#### Generate and write SigProfiler "abstract" 2, 7a, 7b catalogs

We need to adjust the signature names in the exposures
so they match the signature names in `sa.COMPOSITE.sigs`.

```{r SP-abs-syn-cat-COMP}
tmp.exp <- sp.abst.info$syn.exp
rownames(tmp.exp) <- rownames(sa.abst.info$syn.exp)

CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  tmp.exp,
  "sp.sa.COMPOSITE",
  WriteCatCOMPOSITE)
```

```{r SP-abs-syn-cat-96}

CreateAndWriteCatalog(
  sp.sigs, 
  sp.abst.info$syn.exp,
  "sp.sp",
  WriteCatSNS96)

```

```{r}
AddAllScripts()
```


## Check results

Set <tt>unlink = FALSE</tt> if you want to keep the results.
Return values are `"ok"` on success.

```{r}
Diff4SynDataSets("syn.2.7a.7b.bladder.and.melanoma", unlink = TRUE)
Diff4SynDataSets("syn.2.7a.7b.abst", unlink = TRUE)
```

