---
title: "Generate cross-matched SignatureAnalyzer and SigProfiler synthetic data to explore extraction of co-occuring high-burden and low-burden signatures"
author: "Steve Rozen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate cross-matched SignatureAnalyzer and SigProfiler synthetic data to explore extraction of co-occuring high-burden and low-burden signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview and rationale

Checking for interference between low-burden and high-burden signatures,
using MSI-high tumours as the source of high-burden signatures.

## Required libraries

```{r readSPprofiles}
library(ICAMS)
library(SynSig)
```

## Data set 1.1, MSI high tumours from ColoRect-AdenoCA, Uterus-AdenoCA


```{r}
SetNewOutDir("../../syn.MSI.high.and.low.v8/")
```

```{r}

maybe.msi.samples <-
  setdiff(colnames(sa.all.real.exposures),
          colnames(sa.no.hyper.real.exposures))

maybe.msi.colorectal <-
  grep("ColoRect-AdenoCA::",maybe.msi.samples,
       fixed = TRUE, value = TRUE)

maybe.msi.uterus <-
  grep("Uterus-AdenoCA::",maybe.msi.samples,
       fixed = TRUE, value = TRUE)

maybe.msi <- sp.all.real.exposures[ , c(maybe.msi.colorectal, maybe.msi.uterus)]
maybe.msi2 <- maybe.msi[    , maybe.msi["SBS10a", ] < 10]
maybe.msi2 <- maybe.msi2[    , maybe.msi2["SBS14", ] < 10] # SBS14 is Combined POLE and MSI

# Look for SBS44, SBS26; SBS6 is rare

sa.maybe.msi2 <- sa.all.real.exposures[ , colnames(maybe.msi2)]

no.msi.colorectal <-
  grep("ColoRect-AdenoCA::", colnames(sa.no.hyper.real.exposures),
       fixed = TRUE, value = TRUE)

sp.no.msi.colorectal <- sp.all.real.exposures[ , no.msi.colorectal]
mean(sp.no.msi.colorectal["SBS1", ])
mean(sp.no.msi.colorectal["SBS5", ])

no.msi.uterus <-
  grep("Uterus-AdenoCA::", colnames(sa.no.hyper.real.exposures),
       fixed = TRUE, value = TRUE)
sp.no.msi.uterus <- sp.all.real.exposures[ , no.msi.uterus]
mean(sp.no.msi.uterus["SBS1", ])
mean(sp.no.msi.uterus["SBS5", ])



```


# Make sure to use similar signatures to 26 and 44 for sa exposures


```{r}

set.seed(191908)
num.syn.tumors <- 500



bladder.info <- 
  SAAndSPSynDataOneCAType(
    sa.no.hyper.real.exposures,
    sp.no.hyper.real.exposures,
    ca.type = "Bladder-TCC",
    num.syn.tumors,
    file.prefix = "BLADDER")
```

```{r}
undebug(SAAndSPSynDataOneCAType)
skin.info <- 
  SAAndSPSynDataOneCAType(
    sa.all.real.exposures,
    sp.all.real.exposures,
    ca.type = "Skin-Melanoma",
    num.syn.tumors,
    file.prefix = "SKIN")
```

Combine the bladder and skin melanoma synthetic exposures.

```{r} 
# debug(MergeExposures)
sa.bladder.skin.syn.exp <-
  MergeExposures(list(bladder.info$sa.syn.exp, skin.info$sa.syn.exp))

sp.bladder.skin.syn.exp <-
  MergeExposures(list(bladder.info$sp.syn.exp, skin.info$sp.syn.exp))

# We will need the exposures later when evaluating the attributed signatures
WriteExposure(sa.bladder.skin.syn.exp, OutDir("sa.exposure.csv"))
WriteExposure(sp.bladder.skin.syn.exp, OutDir("sp.exposure.csv"))

```

### Generate synthetic catalogs for bladder combined with skin melanoma

#### Create synthetic mutational spectra catalogs based on SignatureAnalyzer attributions

```{r}
CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sa.bladder.skin.syn.exp,
  "sa.sa.COMPOSITE",
  WriteCatCOMPOSITE)


CreateAndWriteCatalog(
  sa.96.sigs,
  sa.bladder.skin.syn.exp,
  "sa.sa.96",
  WriteCat96)

```

#### Create synthetic catalogs based on SigProfiler attributions

Create the catalogs. First we need the matching between SigProfiler
and SignatureAnalyzers signatures.

```{r}
sp.sa.map.info <-
  MapSPToSASignatureNamesInExposure(sp.bladder.skin.syn.exp)

CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sp.sa.map.info$exp2,
  "sp.sa.COMPOSITE",
  WriteCatCOMPOSITE)
```

```{r}
knitr::kable(
  sp.sa.map.info$sp.to.sa.sig.match, 
  caption = 'Best matches from SP signatures to SA signatures',
  digits = 4)
```
```{r, echo = F}
if (FALSE) {
knitr::kable(
  sp.sa.map.info$sa.to.sp.sig.match, 
  caption =
    paste('Best (reverse) matches from SA signatures",
          "to SP signatures (for sanity checking)'),
  digits = 4)
}
```


```{r}
CreateAndWriteCatalog(
  sp.sigs,
  sp.bladder.skin.syn.exp,
  "sp.sp",
  WriteCat96)

```

## Version 2: "abstract" synthetic data set from only SBS2, SBS7a, and SBS7b 

1000 synthetic tumors composed from SBS2 and SBS7a,b  Mutation load distributions 
drawn from bladder transitional cell carcinoma (SBS2) and from skin melanoma (SBS7a,b).
Most spectra contain both signatures.
The potential interference is between SBS2 (mainly C > T) and SBS7a,b (mainly C > T).

### Pull together the parameters

```{r}
x.sp.parms <-
  cbind(bladder.info$sp.parms[ , "SBS2", drop = FALSE],
        skin.info$sp.parms[ , c("SBS7a", "SBS7b")])

# Find mapping from SBS2, SBS7a, and SBS7b to SignatureAnalyzer signatures
# assigned to these tumor types

MatchSigs1Direction(
  sp.sigs[ , "SBS2", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(bladder.info$sa.syn.exp))])

MatchSigs1Direction(
  sp.sigs[ , "SBS7a", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(skin.info$sa.syn.exp))])

MatchSigs1Direction(
  sp.sigs[ , "SBS7b", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(skin.info$sa.syn.exp))])


# We select BI..SBS2, BI....SBS7a, BI.....SBS7b from
# the SignatureAnalyzer parameters.

x.sa.parms <-
  cbind(bladder.info$sa.parms[ , "BI_COMPOSITE_SBS2_P", drop = FALSE],
        skin.info$sa.parms[ , c("BI_COMPOSITE_SBS7a_S", 
                                "BI_COMPOSITE_SBS7b_S")])

knitr::kable(
  x.sa.parms, 
  caption = 'SignatureAnalyzer parameters',
  digits = 4
)

knitr::kable(
  x.sp.parms, 
  caption = 'SigProfiler parameters',
  digits = 4
)
```

### Generate synthetic exposures

```{r}
SetNewOutDir("../../syn.2.7a.7b.abst.v8/")
set.seed(55)
num.syn.tumors <- 1000
```

```{r SP-abs-syn-exp}
sp.abst.info <-
  GenerateSynAbstract(
    x.sp.parms, num.syn.tumors, "sp", "SP.Syn.Abst")
```

```{r SA-abs-syn-exp}
# debug(present.sigs)
# debug(GenerateSyntheticExposures)
sa.abst.info <-
  GenerateSynAbstract(
    x.sa.parms, num.syn.tumors, "sa", "SA.Syn.Abst")

```

### Generate and write SignatureAnalyzer "abstract" 2, 7a, 7b catalogs

```{r SA-abs-syn-cat}
CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sa.abst.info$syn.exp,
  "sa.sa.COMPOSITE",
  WriteCatCOMPOSITE)

CreateAndWriteCatalog(
  sa.96.sigs, 
  sa.abst.info$syn.exp,
  "sa.sa.96",
  WriteCat96)

```

#### Generate and write SigProfiler "abstract" 3, 5, 40 catalogs

We need to adjust the signature names in the exposures
so they match the signature names in \code{sa.COMPOSITE.sigs}.

```{r SP-abs-syn-cat-COMP}
tmp.exp <- sp.abst.info$syn.exp
rownames(tmp.exp) <- rownames(sa.abst.info$syn.exp)

CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  tmp.exp,
  "sp.sa.COMPOSITE",
  WriteCatCOMPOSITE)
```

```{r SP-abs-syn-cat-96}

CreateAndWriteCatalog(
  sp.sigs, 
  sp.abst.info$syn.exp,
  "sp.sp",
  WriteCat96)

```

