---
title: "Generate cross-matched SignatureAnalyzer and SigProfiler synthetic data to explore handling overlapping / interferring signatures"
author: "Steve Rozen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate cross-matched SignatureAnalyzer and SigProfiler synthetic data to explore handling overlapping / interferring signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview and rationale

500 synthetic bladder transitional cell carcinomas (high in SBS2 and SBS13) and 500 synthetic skin melanomas (high in SBS7a,b,c,d).  The potential interference is between 
SBS2 (mainly C > T) and SBS7a,b (mainly C > T)


## Required libraries

```{r readSPprofiles}
library(ICAMS)
library(SynSig)
```

## Data set 1, Kidney-RCC and ovarian adenocarcinoma to explore SBS3, SBS5, and SBS40

It is easier to manage if the output is
is not in a git-managed directory.
```{r}
SetNewOutDir("../../syn.3.5.40.bladder.and.melanoma.v8/")
```

```{r}
set.seed(191906)
num.syn.tumors <- 500

bladder.info <- 
  SAAndSPSynDataOneCAType(
    sa.no.hyper.real.exposures,
    sp.no.hyper.real.exposures,
    ca.type = "Bladder-TCC",
    num.syn.tumors,
    file.prefix = "BLADDER")
```

```{r}
undebug(SAAndSPSynDataOneCAType)
skin.info <- 
  SAAndSPSynDataOneCAType(
    sa.all.real.exposures,
    sp.all.real.exposures,
    ca.type = "Skin-Melanoma",
    num.syn.tumors,
    file.prefix = "SKIN")
```

Combine the RCC and ovarian adenocarcinoma synthetic exposures.

```{r} 
# debug(MergeExposures)
sa.bladder.skin.syn.exp <-
  MergeExposures(list(bladder.info$sa.syn.exp, skin.info$sa.syn.exp))

sp.bladder.skin.syn.exp <-
  MergeExposures(list(bladder.info$sp.syn.exp, skin.info$sp.syn.exp))

# We will need the exposures later when evaluating the attributed signatures
WriteExposure(sa.bladder.skin.syn.exp, OutDir("sa.exposure.csv"))
WriteExposure(sp.bladder.skin.syn.exp, OutDir("sp.exposure.csv"))

```

### Generate synthetic catalogs for RCC combined with ovary adenocarcinoma

#### Create synthetic mutational spectra catalogs based on SignatureAnalyzer attributions

```{r}
CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sa.bladder.skin.syn.exp,
  "sa.sa.COMPOSITE",
  WriteCatCOMPOSITE)


CreateAndWriteCatalog(
  sa.96.sigs,
  sa.bladder.skin.syn.exp,
  "sa.sa.96",
  WriteCat96)

```

#### Create synthetic catalogs based on SigProfiler attributions

Create the catalogs. First we need the matching between SigProfiler
and SignatureAnalyzers signatures.

```{r}
sp.sa.map.info <-
  MapSPToSASignatureNamesInExposure(sp.bladder.skin.syn.exp)

CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sp.sa.map.info$exp2,
  "sp.sa.COMPOSITE",
  WriteCatCOMPOSITE)
```

```{r}
knitr::kable(
  sp.sa.map.info$sp.to.sa.sig.match, 
  caption = 'Best matches from SP signatures to SA signatures',
  digits = 4)
```
```{r}

knitr::kable(
  sp.sa.map.info$sa.to.sp.sig.match, 
  caption =
    paste('Best (reverse) matches from SA signatures",
          "to SP signatures (for sanity checking)'),
  digits = 4)

```


```{r}
CreateAndWriteCatalog(
  sp.sigs,
  sp.bladder.skin.syn.exp,
  "sp.sp",
  WriteCat96)

```

## Version 2: Pull out only SBS2, SBS7a, and SBS7band make an "abstract" synthetic data set

1000 synthetic tumours composed from SBS2 and SBS7a,b  Mutation load distributions drawn from bladder transitional cell carcinoma (SBS2) and from skin melanoma (SBS7a,b).  Most spectra contain both signatures.
The potential interference is between SBS2 (mainly C > T) and SBS7a,b (mainly C > T).


### Pull together the parameters

```{r}
x.sp.parms <-
  cbind(bladder.info$sp.parms[ , "SBS2", drop = FALSE],
        skin.info$sp.parms[ , c("SBS7a", "SBS7b")])

# Find mapping from SBS2, SBS7a, and SBS7b to SignatureAnalyzer signatures
# assigned to these tumor types

MatchSigs1Direction(
  sp.sigs[ , "SBS2", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(bladder.info$sa.syn.exp))])

MatchSigs1Direction(
  sp.sigs[ , "SBS7a", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(skin.info$sa.syn.exp))])

MatchSigs1Direction(
  sp.sigs[ , "SBS7b", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(skin.info$sa.syn.exp))])


# We select BI..SBS2, BI....SBS7a, BI.....SBS7b

x.sa.parms <-
  cbind(bladder.info$sa.parms[ , "BI_COMPOSITE_SBS2_P", drop = FALSE],
        skin.info$sa.parms[ , c("BI_COMPOSITE_SBS7a_S", 
                                "BI_COMPOSITE_SBS7b_S")])

knitr::kable(
  x.sa.parms, 
  caption = 'SignatureAnalyzer parameters',
  digits = 4
)

knitr::kable(
  x.sp.parms, 
  caption = 'SigProfiler parameters',
  digits = 4
)
```

As above, we do not put the output in a git-managed directory.
```{r}
SetNewOutDir("../../syn.2.7a.7b.abst.v8/")
set.seed(44)
num.syn.tumors <- 1000

sp.abst.info <-
  GenerateSynAbstract(
    x.sp.parms, num.syn.tumors, "sp", "SP.Syn.Abst")

sa.abst.info <-
  GenerateSynAbstract(
    x.sa.parms,
    num.syn.tumors, "sa", "SA.Syn.Abst")

```

#### Generate and write SignatureAnalyzer "abstract" 2, 7a, 7b catalogs

```{r}
CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sa.abst.info$syn.exp,
  "sa.sa.COMPOSITE",
  WriteCatCOMPOSITE)

CreateAndWriteCatalog(
  sa.96.sigs, 
  sa.abst.info$syn.exp,
  "sa.sa.96",
  WriteCat96)

```

#### Generate and write SigProfiler "abstract" 3, 5, 40 catalogs

We need to adjust the signature names in the exposures
so they match the signature names in \code{sa.COMPOSITE.sigs}.

```{r}
tmp.exp <- sp.abst.info$syn.exp
rownames(tmp.exp) <- rownames(sa.abst.info$syn.exp)

CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  tmp.exp,
  "sp.sa.COMPOSITE",
  WriteCatCOMPOSITE)

CreateAndWriteCatalog(
  sp.sigs, 
  sp.abst.info$syn.exp,
  "sp.sp",
  WriteCat96)

```

