---
title: "Generate cross-matched SignatureAnalyzer and SigProfiler synthetic data to explore extraction of co-occuring high-burden and low-burden signatures"
author: "Steve Rozen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate cross-matched SignatureAnalyzer and SigProfiler synthetic data to explore extraction of co-occuring high-burden and low-burden signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview and rationale

Checking for interference between low-burden and high-burden signatures,
using MSI-high tumours as the source of high-burden signatures.

## Required libraries

```{r readSPprofiles}
library(ICAMS)
library(SynSig)
```

## Get actual exposures for high-mutational burden signatures, SBS26 and SBS44


```{r}

maybe.msi.samples <-
  setdiff(colnames(sa.all.real.exposures),
          colnames(sa.no.hyper.real.exposures))

maybe.msi.colorectal <-
  grep("ColoRect-AdenoCA::",maybe.msi.samples,
       fixed = TRUE, value = TRUE)

maybe.msi.uterus <-
  grep("Uterus-AdenoCA::",maybe.msi.samples,
       fixed = TRUE, value = TRUE)

maybe.msi <- sp.all.real.exposures[ , c(maybe.msi.colorectal, maybe.msi.uterus)]
maybe.msi2 <- maybe.msi[    , maybe.msi["SBS10a", ] < 10]

 # SBS14 is Combined POLE and MSI, so we do not use it.
sp.msi.exp <- maybe.msi2[    , maybe.msi2["SBS14", ] < 10]
# Look for SBS44, SBS26; SBS6 is rare

sa.msi.exp <- sa.all.real.exposures[ , colnames(sp.msi.exp)]

sp.msi.exp <- sp.msi.exp[c("SBS26", "SBS44"), ]
sa.msi.exp <- sa.msi.exp[c("BI_COMPOSITE_SBS26_S", "BI_COMPOSITE_SBS44_S"), ]

```

## Compute parameters and create synthetic expsoures for high-burden signtures

\code{SAAndSPSynDataOneCAType} will always include at least 1 non-zero exposure
in the the synthetic exposure output.

```{r}

SetNewOutDir("../../syn.MSI.high.with.low.v8/")

set.seed(191908)
num.syn.tumors <- 500

MSI.info <- 
  SAAndSPSynDataOneCAType(
    sa.msi.exp,
    sp.msi.exp,
    ca.type = NULL, # Do not filter by cancer type
    num.syn.tumors,
    file.prefix = "MSI")
```

## TO DO, triple the exposures of the MSI high tumors.

## Get actual exposures for low-mutational burden signatures, SBS1 and SBS2

```{r}

no.msi.colorectal <-
  grep("ColoRect-AdenoCA::", colnames(sa.no.hyper.real.exposures),
       fixed = TRUE, value = TRUE)

sp.no.msi.colorectal <- sp.all.real.exposures[ , no.msi.colorectal]
mean(sp.no.msi.colorectal["SBS1", ])
mean(sp.no.msi.colorectal["SBS5", ])

no.msi.uterus <-
  grep("Uterus-AdenoCA::", colnames(sa.no.hyper.real.exposures),
       fixed = TRUE, value = TRUE)
sp.no.msi.uterus <- sp.all.real.exposures[ , no.msi.uterus]
mean(sp.no.msi.uterus["SBS1", ])
mean(sp.no.msi.uterus["SBS5", ])

no.msi.colnames <- c(no.msi.colorectal, no.msi.uterus)

sp.no.msi.exp <-
  sp.all.real.exposures[c("SBS1", "SBS5"), no.msi.colnames]
stopifnot(dim(sp.no.msi.exp) == c(2, 81)) # sanity check

sa.no.msi.exp1 <-
  sa.all.real.exposures[, no.msi.colnames]
sa.no.msi.exp2 <- sa.no.msi.exp1[rowSums(sa.no.msi.exp1) > 0, ]

sp.sa.map.info <-
  MapSPToSASignatureNamesInExposure(
    sp.exposures = sp.no.msi.exp,
    sa.sig.names.to.consider = rownames(sa.no.msi.exp2))
```

```{r}
knitr::kable(
  sp.sa.map.info$sp.to.sa.sig.match, 
  caption = 'Best matches from SP signatures to SA signatures with exposures in MSI tumors',
  digits = 4)
```



```{r SetSANoMSIExposures}
sa.no.msi.exp3 <- sa.no.msi.exp2[sp.sa.map.info$sp.to.sa.sig.match$to, ]

# View(sa.all.real.exposures[ , no.msi.colnames])
stopifnot(dim(sa.no.msi.exp3) == c(2, 81)) # sanity check

```

## Compute parameters and create synthetic expsoures for low-burden signtures

```{r}

num.syn.tumors <- 500

no.MSI.info <- 
  SAAndSPSynDataOneCAType(
    sa.no.msi.exp3,
    sp.no.msi.exp,
    ca.type = NULL, # Do not filter by cancer type
    num.syn.tumors,
    file.prefix = "MSI")
```

## Exposures for a set of ~500 tumours with SBS1, SBS5, SBS26, and SBS44

Most tumors have a mix of all 4 signatures. This reflects
the SignatureAnalyzer strategy of dealing with the
"hypermutated" tumours separately.

Combine the high and low burden synthetic exposures.

```{r} 
sa.hyper.exp <-
  rbind(MSI.info$sa.syn.exp,
        no.MSI.info$sa.syn.exp)

sp.hyper.exp <-
  rbind(MSI.info$sp.syn.exp,
        no.MSI.info$sp.syn.exp)

```

# NOT DONE

## TO remove "hyper" exposures that contain neither SBS26 nor SBS44

## TO DO catalogs for 500 tumors with neither SBS26 nor SBS44

## TO DO catalogs for ~500 tumor with SBS26 and SBS44

## TO DO Run SignatureAnalyzer first on the 500 tumors with neither SBS26 nor SBS44, and then on the "hyper-mutated tumors"

## TO DO Run SigProfiler on the rbind of the first and second catalogs.

## TO DO maybe Run SigProfiler on the second catalog alone

# FROM HERE DOWN -- OLD

#### Create synthetic mutational spectra catalogs based on SignatureAnalyzer attributions

```{r}
CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sa.bladder.skin.syn.exp,
  "sa.sa.COMPOSITE",
  WriteCatCOMPOSITE)


CreateAndWriteCatalog(
  sa.96.sigs,
  sa.bladder.skin.syn.exp,
  "sa.sa.96",
  WriteCat96)

```

#### Create synthetic catalogs based on SigProfiler attributions

Create the catalogs. First we need the matching between SigProfiler
and SignatureAnalyzers signatures.

```{r}
sp.sa.map.info <-
  MapSPToSASignatureNamesInExposure(sp.bladder.skin.syn.exp)

CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sp.sa.map.info$exp2,
  "sp.sa.COMPOSITE",
  WriteCatCOMPOSITE)
```


```{r}
CreateAndWriteCatalog(
  sp.sigs,
  sp.bladder.skin.syn.exp,
  "sp.sp",
  WriteCat96)

```

## Version 2: "abstract" synthetic data set from only SBS2, SBS7a, and SBS7b 

1000 synthetic tumors composed from SBS2 and SBS7a,b  Mutation load distributions 
drawn from bladder transitional cell carcinoma (SBS2) and from skin melanoma (SBS7a,b).
Most spectra contain both signatures.
The potential interference is between SBS2 (mainly C > T) and SBS7a,b (mainly C > T).

### Pull together the parameters

```{r}
x.sp.parms <-
  cbind(bladder.info$sp.parms[ , "SBS2", drop = FALSE],
        skin.info$sp.parms[ , c("SBS7a", "SBS7b")])

# Find mapping from SBS2, SBS7a, and SBS7b to SignatureAnalyzer signatures
# assigned to these tumor types

MatchSigs1Direction(
  sp.sigs[ , "SBS2", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(bladder.info$sa.syn.exp))])

MatchSigs1Direction(
  sp.sigs[ , "SBS7a", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(skin.info$sa.syn.exp))])

MatchSigs1Direction(
  sp.sigs[ , "SBS7b", drop = F], 
  sa.96.sigs[ , FixSASigNames(rownames(skin.info$sa.syn.exp))])


# We select BI..SBS2, BI....SBS7a, BI.....SBS7b from
# the SignatureAnalyzer parameters.

x.sa.parms <-
  cbind(bladder.info$sa.parms[ , "BI_COMPOSITE_SBS2_P", drop = FALSE],
        skin.info$sa.parms[ , c("BI_COMPOSITE_SBS7a_S", 
                                "BI_COMPOSITE_SBS7b_S")])

knitr::kable(
  x.sa.parms, 
  caption = 'SignatureAnalyzer parameters',
  digits = 4
)

knitr::kable(
  x.sp.parms, 
  caption = 'SigProfiler parameters',
  digits = 4
)
```

### Generate synthetic exposures

```{r}
SetNewOutDir("../../syn.2.7a.7b.abst.v8/")
set.seed(55)
num.syn.tumors <- 1000
```

```{r SP-abs-syn-exp}
sp.abst.info <-
  GenerateSynAbstract(
    x.sp.parms, num.syn.tumors, "sp", "SP.Syn.Abst")
```

```{r SA-abs-syn-exp}
# debug(present.sigs)
# debug(GenerateSyntheticExposures)
sa.abst.info <-
  GenerateSynAbstract(
    x.sa.parms, num.syn.tumors, "sa", "SA.Syn.Abst")

```

### Generate and write SignatureAnalyzer "abstract" 2, 7a, 7b catalogs

```{r SA-abs-syn-cat}
CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sa.abst.info$syn.exp,
  "sa.sa.COMPOSITE",
  WriteCatCOMPOSITE)

CreateAndWriteCatalog(
  sa.96.sigs, 
  sa.abst.info$syn.exp,
  "sa.sa.96",
  WriteCat96)

```

#### Generate and write SigProfiler "abstract" 3, 5, 40 catalogs

We need to adjust the signature names in the exposures
so they match the signature names in \code{sa.COMPOSITE.sigs}.

```{r SP-abs-syn-cat-COMP}
tmp.exp <- sp.abst.info$syn.exp
rownames(tmp.exp) <- rownames(sa.abst.info$syn.exp)

CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  tmp.exp,
  "sp.sa.COMPOSITE",
  WriteCatCOMPOSITE)
```

```{r SP-abs-syn-cat-96}

CreateAndWriteCatalog(
  sp.sigs, 
  sp.abst.info$syn.exp,
  "sp.sp",
  WriteCat96)

```

