---
title: "Generate cross-matched SignatureAnalyzer and SigProfiler synthetic data similar to pancreatic adenocarcinomas"
author: "Steve Rozen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate cross-matched SignatureAnalyzer and SigProfiler synthetic data similar to pancreatic adenocarcinomas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Required libraries and set ouput path

```{r librariesEtc}
library(ICAMS)
library(SynSig)
SetNewOutDir("../../syn.pancreas.v8/")
```

## Generate synthetic exposures

```{r GenExposures}
set.seed(191907)
num.syn.tumors <- 1000

panc.info <- 
  SAAndSPSynDataOneCAType(
    sa.no.hyper.real.exposures,
    sp.no.hyper.real.exposures,
    ca.type = "Panc-AdenoCA",
    num.syn.tumors,
    file.prefix = "PANC")
```

#### Create synthetic mutational spectra catalogs based on SignatureAnalyzer attributions

```{r}
CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  panc.info$sa.syn.exp,
  "sa.sa.COMPOSITE",
  WriteCatCOMPOSITE)


CreateAndWriteCatalog(
  sa.96.sigs,
  panc.info$sa.syn.exp,
  "sa.sa.96",
  WriteCat96)

```

#### Create synthetic catalogs based on SigProfiler attributions

First we need the matching between SigProfiler
and SignatureAnalyzers signatures.

```{r}
sp.sa.map.info <-
  MapSPToSASignatureNamesInExposure(panc.info$sp.syn.exp)

CreateAndWriteCatalog(
  sa.COMPOSITE.sigs,
  sp.sa.map.info$exp2,
  "sp.sa.COMPOSITE",
  WriteCatCOMPOSITE)
```

### The mapping from SP signatures to SA signatures

```{r}
knitr::kable(
  sp.sa.map.info$sp.to.sa.sig.match, 
  caption = 'Best matches from SP signatures to SA signatures',
  digits = 4
)
```
```{r echo=FALSE}
if (FALSE) {
knitr::kable(
  sp.sa.map.info$sa.to.sp.sig.match, 
  caption =
    paste('Best (reverse) matches from SA signatures",
          "to SP signatures (for sanity checking)'),
  digits = 4
)
}
```


```{r}
CreateAndWriteCatalog(
  sp.sigs,
  panc.info$sp.syn.exp,
  "sp.sp",
  WriteCat96)

```

## Session Info

```{r}
sessionInfo()
```
